Cognitive Analysis

### 2024-10-18 Manuscript Analysis/manuscript_analysis.Rmd
reran initial analyses

@@

# Color Palette for Figures
my_pal <- c("#033F63", "#2CB5AE", "#28666E", "#4BAA9B", "#49E9B9", "#57A773")

@@

@@

@@

rm(cog_table, demo_table, demo)

@@

m0_predicted_values |>
  ggplot() +
  aes(x = time_norm,
      y = fit,
      color = speaker,
      fill = speaker) +
  geom_ribbon(aes(ymin = ul, ymax = ll), alpha = .5, color = NA) +
  geom_line() +
  geom_vline(xintercept = 0, size = 0.35) +
  geom_vline(xintercept = 1910, linetype = 2) +
  annotate("rect", xmin = 500, xmax = 2910, ymin = 0, ymax = 200, alpha = 0.15) +
  coord_cartesian(ylim = c(-10, 250), xlim = c(0, 4910)) +
  facet_wrap("rep_acc") +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Normalized Time (ms)",
       y = "Pupil Dilation (Arbitrary Units)",
       color = "Speaker",
       fill = "Speaker",
       linetype = "Speaker") +
  scale_color_manual(values = c(my_pal[1], my_pal[5]), labels = c("Neurotypical", "ALS")) +
  scale_fill_manual(values = c(my_pal[1], my_pal[5]), labels = c("Neurotypical", "ALS"))

@@

@@

wm_predicted_values |>
  ggplot() +
  aes(x = list_sort_c,
      y = fit,
      color = rep_acc,
      fill = rep_acc) +
  geom_ribbon(aes(ymin = ul, ymax = ll), alpha = .5, color = NA) +
  geom_line() +
  facet_wrap("speaker") +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Working Memory",
       y = "Pupil Dilation (Arbitrary Units)",
       color = "Repetition Accuracy",
       fill = "Repetition Accuracy",
       linetype = "Repetition Accuracy") +
  scale_color_manual(values = c(my_pal[1], my_pal[2])) +
  scale_fill_manual(values = c(my_pal[1], my_pal[2]))

@@

@@

inhib_predicted_values |>
  ggplot() +
  aes(x = flanker_c,
      y = fit,
      color = rep_acc,
      fill = rep_acc) +
  geom_ribbon(aes(ymin = ul, ymax = ll), alpha = .5, color = NA) +
  geom_line() +
  facet_wrap("speaker") +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Inhibitory Control",
       y = "Pupil Dilation (Arbitrary Units)",
       color = "Repetition Accuracy",
       fill = "Repetition Accuracy",
       linetype = "Repetition Accuracy") +
  scale_color_manual(values = c(my_pal[1], my_pal[2])) +
  scale_fill_manual(values = c(my_pal[1], my_pal[2]))

@@

@@

flex_predicted_values |>
  ggplot() +
  aes(x = card_sort_c,
      y = fit,
      color = rep_acc,
      fill = rep_acc) +
  geom_ribbon(aes(ymin = ul, ymax = ll), alpha = .5, color = NA) +
  geom_line() +
  facet_wrap("speaker") +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Cognitive Flexibility",
       y = "Pupil Dilation (Arbitrary Units)",
       color = "Repetition Accuracy",
       fill = "Repetition Accuracy",
       linetype = "Repetition Accuracy") +
  scale_color_manual(values = c(my_pal[1], my_pal[2])) +
  scale_fill_manual(values = c(my_pal[1], my_pal[2]))

@@

@@

process_predicted_values |>
  ggplot() +
  aes(x = pattern_c,
      y = fit,
      color = rep_acc,
      fill = rep_acc) +
  geom_ribbon(aes(ymin = ul, ymax = ll), alpha = .5, color = NA) +
  geom_line() +
  facet_wrap("speaker") +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Processing Speed",
       y = "Pupil Dilation (Arbitrary Units)",
       color = "Repetition Accuracy",
       fill = "Repetition Accuracy",
       linetype = "Repetition Accuracy") +
  scale_color_manual(values = c(my_pal[1], my_pal[2])) +
  scale_fill_manual(values = c(my_pal[1], my_pal[2]))

@@

m17 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + 
                 speaker*card_sort_c+ speaker*list_sort_c + 
                 rep_acc*list_sort_c + rep_acc*card_sort_c +
                 (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m17,
                  dv.labels = "Perceived Listening Effort")

@@

@@

ggeffects::ggpredict(m17, terms = c("list_sort_c", "speaker", "rep_acc")) |> 
  plot()

>>>>

### Distribution of Cognitive Scores

@@

# Color Palette for Figures
my_pal <- c("#25CED1", "#0D3B66", "#F75C03", "#3B5249", "#F1C40F", "#D90368")

@@

ggsave("Figures\\cog_distribution.png", plot = last_plot())

```
### Correlation Among Cognitive Subtests

ggsave("Figures\\cog_correlation.png", plot = last_plot())

@@

```{r, echo = F}

@@

rm(cog_table, demo)

@@

m0_predicted_values |>
  ggplot() +
  aes(x = time_norm,
      y = fit,
      color = speaker,
      linetype = rep_acc,
      fill = speaker) +
  geom_line(size = .8) +
  geom_vline(xintercept = 0, size = 0.35) +
  geom_vline(xintercept = 1910, linetype = 2) +
  annotate("rect", xmin = 500, xmax = 2910, ymin = 0, ymax = 200, alpha = 0.15) +
  coord_cartesian(ylim = c(-10, 250), xlim = c(0, 4910)) +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Normalized Time (ms)",
       y = "Pupil Dilation (Arbitrary Units)",
       color = "Speaker",
       linetype = "Repetition Accuracy") +
  scale_color_manual(values = c(my_pal[1], my_pal[2]), labels = c("Neurotypical", "ALS"))

@@

Overall, the plot is showing that as working memory increases, so does pupil dilaiton. However, there are some interesting trends in the data.

```{r, message = F, warning = F}

@@

wm_predicted_values |>
  ggplot() +
  aes(x = list_sort_c,
      y = fit,
      color = speaker,
      linetype = rep_acc) +
  geom_line(size = .8) +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Working Memory",
       y = "Pupil Dilation (Arbitrary Units)",
       color = "Speaker",
       linetype = "Repetition Accuracy") +
  scale_color_manual(values = c(my_pal[1], my_pal[2]))

@@

Overall, there seems to be differing trends between the relationship between inhibitory control and pupil dilation depending on the speaker and phrase repetition accuracy. For the accurately recognized phrases from the control speaker, it appears that as inhibitory control increases, pupil dilation also increases. However, for the other three conditions, it seems there is a slight quadratic relationship, where listeners with lower and higher inhibitory control both have larger pupil dilation, but listeners with inhibitory scores around 70-85 (i.e., 1-2 SD below average standardized score 100), have the lowest pupil dilation.

@@

inhib_predicted_values |>
  ggplot() +
  aes(x = flanker_c,
      y = fit,
      color = speaker,
      linetype = rep_acc) +
  geom_line(size = .8) +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Inhibitory Control",
       y = "Pupil Dilation (Arbitrary Units)",
       color = "Speaker",
       linetype = "Repetition Accuracy") +
  scale_color_manual(values = c(my_pal[1], my_pal[2]))
```

@@

General trend is that as cognitive flexibility increases, so does pupil dilation. However, there is likely an effect of phrase repetition accuracy. Pupil dilation is higher for inaccurately recognized phrases (for both speakers, but especially the ALS speaker) for listeners with cognitive flexibility scores over 105.

@@

flex_predicted_values |>
  ggplot() +
  aes(x = card_sort_c,
      y = fit,
      color = speaker,
      linetype = rep_acc) +
  geom_line(size = .8) +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Cognitive Flexibility",
       y = "Pupil Dilation (Arbitrary Units)",
       color = "Speaker",
       linetype = "Repetition Accuracy") +
  scale_color_manual(values = c(my_pal[1], my_pal[2]))
```

@@

For Processing speed, there is a curvilinear relationship. Listeners with lower and higher processing speed scores tend to have higher pupil dilation, especially for ALS speaker (both accurately and inaccurately recognized phrases), and accurately recognized control phrases.

@@

process_predicted_values |>
  ggplot() +
  aes(x = pattern_c,
      y = fit,
      color = speaker,
      linetype = rep_acc) +
  geom_line(size = .8) +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Processing Speed",
       y = "Pupil Dilation (Arbitrary Units)",
       color = "Speaker",
       linetype = "Repetition Accuracy") +
  scale_color_manual(values = c(my_pal[1], my_pal[2]))

@@

m17 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + 
                 speaker*scale(card_sort_c, scale = F) + speaker*scale(list_sort_c, scale = F) + 
                 rep_acc*scale(list_sort_c, scale = F) + rep_acc*scale(card_sort_c, scale = F) +
                 (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m17,
                  dv.labels = "Perceived Listening Effort")

@@

The final best-fitting LME model for the PLE data is below. In general, both the interaction between working memory and cognitive flexibility with speaker was significant, indicating that listeners with higher working memory scores generally rated accurately-recognized ALS speaker as more effortful. However, listener with higher cognitive flexibility scores tend to rate accurately-recognized ALS speaker as less effortful.The three-way interaction for both of these cognitive measures with speaker and repetition accuracy was not significant.

These cognitive measures also had a significant interaction with repetition accuracy. Listeners with higher working memory scores tend to rate inaccurately recognized phrases from the control speaker as less effortful. Additionally, listeners with higher cognitive flexibility scores tend to rate inaccurately-recognized phrases for the control speaker as more effortful. As mentioned before, the three way interactions were not significant indicating that the pattern of results for inaccurately recognized ALS phrases was similar.

@@
